Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Chinook Salmon Populations

### 2022-01-31 analysis/paper/paper.Rmd
incorporated TB results edits

@@

<!--
Lastly, we explored one additional multiple-census estimator, the Schumacher-Eschmeyer, which is based on minimizing the weighted sum of squares between the proportion of marked individuals in the sample and the unknown proportion of marked individuals in the population. The Schumacher-Eschmeyer estimator is as follows:

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M^2_i}{\sum\limits_{i = 1}^k m_i M_i}
$$
-->

@@

Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 3,000 J/g for invertebrates and 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. Because were were unable to differentiate juvenile Chinook Salmon from other fish prey, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. The average length of Northern Pikeminnow caught in Deadwater Slough during our study was converted to weight using a weight-length formula from @Parker1995. This resulted in average spring and fall starting eights of 504.4 and 670.5 g for Northern Pikeminnow with average lengths of `r avg_fall_NPM` and `r avg_spring_NPM` mm, respectively. We assumed no growth in individual Northern Pikeminnow, resulting in equal start and end weights. Mean daily water temperatures were summarized from 15-minute interval temperature readings between March 3, 2013 and June 14, 2021 from USGS gage station 13307000, approximately 22 RKM downstream of Deadwater Slough.

@@

Mark-recapture abundance estimates for Northern Pikeminnow in Deadwater Slough during fall 2019 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` and for fall 2020 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` (Table \@ref(tab:abund-tab)).For both estimates, the multiple census estimates were larger than the single census estimates and small numbers of recaptures resulted in wide confidence intervals (***figure reference***).

All mark-recapture estimators yielded estimates with overlapping confidence intervals. However, because our sampling design most closely matched a multiple census estimator (i.e., we released marked fish each day and recaptures from all previous days were recorded for each day) and required a small sample size modification, we used the Schnabel estimates of abundance for further analyses and results. Accordingly, the mean of Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_F, spring_abund$Uci_F)), big.mark = ","), collapse = " - ")`). By multiplying the ratio (`r round(total_cpue$catch_per_hr[total_cpue$event_name == "Spring 2021"] / mean(total_cpue$catch_per_hr[total_cpue$event_name %like% "Fall"]), 2)`) between the spring CPUE (`r round(total_cpue$catch_per_hr[total_cpue$event_name == "Spring 2021"], 2)`) and mean of the fall CPUEs (`r round(mean(total_cpue$catch_per_hr[total_cpue$event_name %like% "Fall"]), 2)`), we estimated there to be `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough during spring 2021.

<!--However, because our sampling design most closely matched a multiple census estimator (i.e., we released marked fish each day and recaptures from all previous days were recorded for each day), we fell those estimators most appropriate. Additionally, because the Schnabel estimator includes a small sample size modification, as in our case (i.e., few recaptures), we feel it is most appropriate. The Schnabel estimator has the added benefit of more precise confidence intervals and is more straightforward than the Schumacher-Eschmeyer estimate as it is essentially the weighted average among multiple Chapman-modified Lincoln-Peterson estimates. Therefore, further analyses and results will be based on the Schnabel estimates of abundance.-->

The PSD for Northern Pikeminnow in Deadwater Slough across all three efforts was `r prettyNum(psd, digits = 1)`% i.e., `r prettyNum(psd, digits = 1)`% of captured fish were over the quality size classification defined by @Winther2020. For comparison, @Winther2020 reported a PSD for Northern Pikeminnow captured below Bonneville Dam of 41% and 18% in Bonneville Reservoir, suggesting that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. In fact, the state record Northern Pikeminnow, which measured `r max(cmr_df$length, na.rm = T)` mm (`r round(max(cmr_df$length, na.rm = T) / 25.4, 1)` in) TL, was caught in Deadwater Slough during the fall 2020 effort.

@@

We examined the stomach contents of Northern Pikeminnow from Deadwater Slough using gastric lavage. To validate the efficacy of the method, we euthanized 75 Northern Pikeminnow after gastric lavage was completed and removed the remaining food items via dissection. We found that gastric lavage successfully removed food items in all dissected individuals, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in some form in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) of those. Northern Pikeminnow captured during the spring 2021 effort had a higher proportion of diets with fish contents (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Spring_2021"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Spring_2021"], 1)`%) than fish captured in the fall 2019 (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Fall_2019"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Fall_2019"], 1)`%) and fall 2020 (`r round(100 * gastric_tbl$fish_contents[gastric_tbl$event_name == "Fall_2020"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Fall_2020"], 1)`%) (Table \@ref(tab:gastric-tab)). For those individuals with stomach contents, the average wet weight of total contents was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; range = `r round(min(stmch_df$stomach_contents_weight), 2)`-`r round(max(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants were `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall stomach contents examined.

@@

We estimated an individual, average-size (`r avg_fall_NPM` mm) Northern Pikeminnow to consume `r median(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` g (range: `r min(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` - `r max(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` g) of fish during the DSR emigration to maintain their body size, depending on the proportion of diet consisting of fish versus invertebrates (Figure \@ref(fig:bio-p)). Similarly, we estimated an average-size (`r avg_spring_NPM` mm) Northern Pikeminnow to consume `r median(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` g (range: `r min(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` - `r max(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` g) of fish during the NRR emigration (Figure \@ref(fig:bio-p)). Given a mean Northern Pikeminnow population size of `r prettyNum(unique(impact_df$N[impact_df$season == "Fall"]), big.mark = ",")` from fall estimates and `r prettyNum(unique(impact_df$N[impact_df$season == "Spring"]), big.mark = ",")` for spring 2021 and using the median value from diet scenarios (i.e., 50% of their diet consists of fish), we'd estimate that the Northern Pikeminnow population can consume `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 50]), big.mark = ",")` g (range: `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 10]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 10]), big.mark = ",")` - `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 90]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 90]), big.mark = ",")` g) fish during the DSR emigration and an additional `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 50]), big.mark = ",")` g (range: `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 10]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 10]), big.mark = ",")` - `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 90]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 90]), big.mark = ",")` g) during the NRR emigration in Deadwater Slough. In total, the Northern Pikeminnow population can consume `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 50]) + unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 50]), big.mark = ",")` g of fish during the DSR and NRR emigrations to maintain their population size if we assume 50% of their diet consists of fish.

@@

1. Start with total g of fish consumed by Northern Pikemnnow population, summing DSR and NRR time periods.
2. Parse out portion of fish total g consumed that is juvenile Chinook Salmon (1%, 10%, 30%, 50%).
3. Divide by average weight of juvenile Chinook Salmon (10.6 g), the average of DSR (10.3 g) and NRR (10.9 g) emigrants.
4. Estimate of total juvenile Chinook Salmon consumed.
5. Multiply by SAR to estimate "adult equivalents".

<!--
## Juvenile Chinook Salmon Abundance
For context, throughout the migration periods of DSR and NRR life histories it is estimated that 113,976 individuals are produced and therefore available within the upper Salmon River basin (Table x).

The abundance of juvenile Chinook Salmon for the upper Salmon River basin was estimated at 6 rotary screw traps [@McClure2020]. Overall, in 2020, there were an estimated 59,289 DSR, 22,264 NRR, and 10,159 parr estimated among all these traps in the upper Salmon River basin (Table X.)
-->

>>>>

$$
\hat{N} = \frac{\sum\limits_{i = 1}^k n_i M_{i-1}}{\left(\sum\limits_{i = 1}^k m_i \right) + 1}
$$

@@

@@

Predator energy density for Northern Pikeminnow was fixed at 6,703 Joules(J)/g [@Deslauriers2017]. Prey energy densities were fixed at 3,000 J/g for invertebrates and 21,500 J/g for juvenile Chinook Salmon [@Moss2016]. Because were were unable to differentiate juvenile Chinook Salmon from other fish prey, we assume all fish prey have the same energy densities as juvenile Chinook Salmon. The average length of Northern Pikeminnow caught in Deadwater Slough during our study was converted to weight using a weight-length formula from @Parker1995.

$$
log_{10} W_s = -4.886 + 2.986[log_{10}(FL)]
$$

This resulted in average spring and fall starting eights of 504.4 and 670.5 g for Northern Pikeminnow with average lengths of `r avg_fall_NPM` and `r avg_spring_NPM` mm, respectively. We assumed no growth in individual Northern Pikeminnow, resulting in equal start and end weights. Mean daily water temperatures were summarized from 15-minute interval temperature readings between March 3, 2013 and June 14, 2021 from USGS gage station 13307000, approximately 22 RKM downstream of Deadwater Slough.

@@

Mark-recapture abundance estimates for Northern Pikeminnow in Deadwater Slough during fall 2019 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` and for fall 2020 ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` (Table \@ref(tab:abund-tab)).For both estimates, the multiple census estimates were larger than the single census estimates and small numbers of recaptures resulted in wide confidence intervals (***figure reference***). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, the Schnabel estimator was determined to provide the best estimates of abundance. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_F, spring_abund$Uci_F)), big.mark = ","), collapse = " - ")`).<!--Need to add the delayed-mixing Schnabel estimate results--> For spring 2021, we estimated abundance by multiplying the mean fall abundance by the ratio between the spring CPUE and mean of the fall CPUEs (`r round(total_cpue$catch_per_hr[total_cpue$event_name == "Spring 2021"] / mean(total_cpue$catch_per_hr[total_cpue$event_name %like% "Fall"]), 2)`), resulting in an estimate of `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of ***value*** and ***value*** Northern Pikeminnow per RKM.

<!-- need to adjust the spring estimate -->

The lengths of Northern Pikeminnow captured ranged from ***value*** to ***value*** mm with an average length of ***value*** mm (\@ref(fig:length-freq-p)). The PSD for Northern Pikeminnow in Deadwater Slough across all three efforts was `r prettyNum(psd, digits = 1)`%.

@@

We found that gastric lavage successfully removed food items in all dissected individuals, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs. Northern Pikeminnow captured during spring 2021 had a higher proportion of fish content in their diet compared to individuals captured in the fall surveys (Table \@ref(tab:gastric-tab)). The average wet weight of total contents for an individual was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; range = `r round(min(stmch_df$stomach_contents_weight), 2)`-`r round(max(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants made up `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of all stomach contents examined.

@@

During the fall DSR emigration, we estimated an average-size Northern Pikeminnow (***value*** mm) to consume ***value*** g of fish to maintain their body size, assuming 60% of their diet consists of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from ***value*** g for a diet of 30% fish to ***value*** g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (***value***) to consume ***value*** 44.29 g of fish, assuming 60% of their diet is fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from ***value*** g for a diet of 30% fish to ***value*** g for a diet of 90% fish. Given estimated mean Northern Pikeminnow population sizes of ***value*** from fall estimates and ***value*** for spring 2021, and assuming 60% of their diet is fish, we estimate that Northern Pikeminnow can consume ***value*** metric tons of fish during the fall DSR emigration and an additional ***value*** metric tons during the spring NRR emigration in Deadwater Slough. In total, the Northern Pikeminnow population can consume ***value*** metric tons of fish during the DSR and NRR emigrations to maintain their population size, assuming 60% of their diet consists of fish.

<!--
We estimated an individual, average-size (`r avg_fall_NPM` mm) Northern Pikeminnow to consume `r median(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` g (range: `r min(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` - `r max(bio_g_summ$cumm_g[bio_g_summ$season == "Fall"])` g) of fish during the DSR emigration to maintain their body size, depending on the proportion of diet consisting of fish versus invertebrates (Figure \@ref(fig:bio-p)). Similarly, we estimated an average-size (`r avg_spring_NPM` mm) Northern Pikeminnow to consume `r median(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` g (range: `r min(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` - `r max(bio_g_summ$cumm_g[bio_g_summ$season == "Spring"])` g) of fish during the NRR emigration (Figure \@ref(fig:bio-p)). Given a mean Northern Pikeminnow population size of `r prettyNum(unique(impact_df$N[impact_df$season == "Fall"]), big.mark = ",")` from fall estimates and `r prettyNum(unique(impact_df$N[impact_df$season == "Spring"]), big.mark = ",")` for spring 2021 and using the median value from diet scenarios (i.e., 50% of their diet consists of fish), we'd estimate that the Northern Pikeminnow population can consume `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 50]), big.mark = ",")` g (range: `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 10]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 10]), big.mark = ",")` - `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 90]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 90]), big.mark = ",")` g) fish during the DSR emigration and an additional `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 50]), big.mark = ",")` g (range: `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 10]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 10]), big.mark = ",")` - `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 90]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 90]), big.mark = ",")` g) during the NRR emigration in Deadwater Slough. In total, the Northern Pikeminnow population can consume `r prettyNum(unique(impact_df$g_fish_cons[impact_df$season == "Fall" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Fall" & impact_df$perc_diet == 50]) + unique(impact_df$g_fish_cons[impact_df$season == "Spring" & impact_df$perc_diet == 50]) * unique(impact_df$N[impact_df$season == "Spring" & impact_df$perc_diet == 50]), big.mark = ",")` g of fish during the DSR and NRR emigrations to maintain their population size if we assume 50% of their diet consists of fish.
-->

@@

Using the total g of fish consumed by the Northern Pikeminnow population (summing fall and spring surveys) we produced a sensitivity analysis for three Northern Pikeminnow diet scenarios where juvenile Chinook Salmn make up 30%, 50%, and 65% of total fish prey consumed (Figure ***reference***). Assuming that 50% of the total fish consumed are juvenile Chinook Salmon, we estimate between (30% fish scenario with CIs) and (90% fish scenario with CIs) juvenile Chinook Salmon will be consumed. We then used our estimates of total juvenile Chinook Salmon consumed to estimate "adult equivalents" using the Granite-to-Granite smolt-to-adult survival rate of ***value*** (SE: ***value***; Table ***reference***). Assuming predation in Deadwater Slough was reduced or eliminated, the juvenile consumption would be equivalent to ***value*** adults in the spring (range ***value*** - ***value***) and ***value*** adults (range ***value*** - ***value***) in the fall, totaling ***value*** adult Chinook Salmon.
